---
title: 'Base R: 1D Data'
author: "Jae Yeon Kim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
editor_options: 
  markdown: 
    wrap: 72
---

# Base R: 1D Data

## Setup

-   Check your `dplyr` package is up-to-date by typing
    `packageVersion("dplyr")`. If the current installed version is less
    than 1.0, then update by typing `update.packages("dplyr")`. You may
    need to restart R to make it work.

```{r}
ifelse(packageVersion("dplyr") >= 1,
  "The installed version of dplyr package is greater than or equal to 1.0.0", update.packages("dplyr")
)

if (!require("pacman")) install.packages("pacman")

pacman::p_load(
  tidyverse)
```

## Base R data structure

The rest of the chapter follows the basic structure in [the Data
Wrangling Cheat
Sheet](https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
created by RStudio.

To make the best use of the R language, you'll need a strong
understanding of the basic data types and data structures and how to
operate on those. R is an **object-oriented** language, so the
importance of this cannot be understated.

It is **critical** to understand because these are the objects you will
manipulate on a day-to-day basis in R, and they are not always as easy
to work with as they sound at the outset. Dealing with object
conversions is one of the most common sources of frustration for
beginners.

> To understand computations in R, two slogans are helpful: - Everything
> that exists is an object. - Everything that happens is a function
> call.

> \_\_John Chambers\_\_the creator of S (the mother of R)

1.  [Main Classes](#main-classes) introduces you to R's one-dimensional
    or atomic classes and data structures. R has five basic atomic
    classes: logical, integer, numeric, complex, character. Social
    scientists don't use complex classes.

2.  [Attributes](#attributes) takes a small detour to discuss
    attributes, R's flexible metadata specification. Here, you'll learn
    about factors, an important data structure created by setting
    attributes of an atomic vector. R has many data structures: vector,
    list, matrix, data frame, factors, tables.

![Concept map for data types. By Meghan Sposato, Brendan Cullen, Monica
Alonso.](https://github.com/rstudio/concept-maps/raw/master/en/data-types.svg)

## Atomic classes

`R`'s main atomic classes are:

-   character (or a "string" in Python and Stata)
-   numeric (integer or float)
-   integer (just integer)
-   logical (booleans)

```{r}
chr <- c("a", "b", "c")
num <- c(1L, 2L, 3L) 
dob <- c(1.2, 2.3, 3.4)
log <- c(TRUE, FALSE)

```

| Example                                     | Type      |
|---------------------------------------------|-----------|
| "a", "swc"                                  | character |
| 2, 15.5                                     | numeric   |
| 2 (Must add a `L` at end to denote integer) | integer   |
| `TRUE`, `FALSE`                             | logical   |

Like Python, R is dynamically typed. There are a few differences in
terminology, however, that are pertinent.

-   First, "types" in Python are referred to as "classes" in R.

What is a class?

![from
<https://brilliant.org/>](https://ds055uzetaobb.cloudfront.net/brioche/uploads/pJZt3mh3Ht-prettycars.png?width=2400)

-   Second, R has different names for the types string, integer, and
    float --- specifically **character**, **integer** (not different),
    and **numeric**. Because there is no "float" class in R, users tend
    to default to the "numeric" class when working with numerical data.

The function for recovering object classes is `class()`. L suffix to
qualify any number to make it an explicit integer. See more from the [R
language
definition](https://cran.r-project.org/doc/manuals/R-lang.html).

```{r}
class(3)
class(3L)
class("Three")
class(F)
```

## Data structures

R's base data structures can be organized by their dimensionality (1d,
2d, or nd) and whether they're homogeneous (all contents must be of the
same type) or heterogeneous (the contents can be of different types).
This gives rise to the five data types most often used in data analysis:

|     | Homogeneous   | Heterogeneous |
|-----|---------------|---------------|
| 1d  | Atomic vector | List          |
| 2d  | Matrix        | Data frame    |
| nd  | Array         |               |

Each data structure has its specifications and behavior. For our
purposes, an important thing to remember is that R is always **faster**
(more efficient) working with homogeneous (**vectorized**) data.

## Vector properties

Vectors have three common properties:

-   Class, `class()`, or what type of object it is (same as `type()` in
    Python).
-   Length, `length()`, how many elements it contains (same as `len()`
    in Python).
-   Attributes, `attributes()`, additional arbitrary metadata.

They differ in the types of their elements: all atomic vector elements
must be the same type, whereas the elements of a list can have different
types.

## Creating different types of atomic vectors

Remember, there are four common types of vectors: \* `logical` \*
`integer` \* `numeric` (same as `double`) \* `character`.

You can create an empty vector with `vector()` (By default, the mode is
`logical.` You can be more explicit as shown in the examples below.) It
is more common to use direct constructors such as `character()`,
`numeric()`, etc.

```{r}
x <- vector()

# with a length and type
vector("character", length = 10)

## character vector of length 5
character(5)

numeric(5)

logical(5)
```

Atomic vectors are usually created with `c()`, which is short for
concatenate:

```{r}
x <- c(1:1000)

x

length(x)
```

`x` is a numeric vector. These are the most common kind. You can also
have logical vectors.

```{r}
y <- c(TRUE, TRUE, FALSE, FALSE)

y
```

Finally, you can have character vectors:

```{r}
social_sciences <- c("polisci", "econ", "sociology")

class(social_sciences)

is.integer(social_sciences) # integer?

is.character(social_sciences) # character?

is.atomic(social_sciences) # atomic?

typeof(social_sciences) # what's the type?
```

**Short exercise: Create and examine your vector**

Create a character vector called `fruit` containing 4 of your favorite
fruits. Then evaluate its structure using the commands below.

```{r, eval = FALSE}
# First, create your fruit vector
# YOUR CODE HERE
fruit <- c("pine", "mang", "waterm", "appl")
  # Examine your vector

length(fruit)
class(fruit)
str(fruit)
```

**Add elements**

You can add elements to the end of a vector by passing the original
vector into the `c` function, like the following:

```{r}
z <- c("BTS", "twice", "stella_jang")

z <- c(z, "yiruma")

print(z)
```

```{r}
x <- c(1:3)
```

More examples of vectors

```{r}
x <- c(0.5, 0.7)

x <- c(TRUE, FALSE)

x <- c("a", "b", "c", "d", "e")

x <- 9:100
```

You can also create vectors as a sequence of numbers:

```{r}
seq(10)

seq(1, 10, by = 0.1)
```

Atomic vectors are always flat, even if you nest `c()`'s:

```{r eval = TRUE}
c(1, c(2, c(3, 4)))

# the same as
c(1, 2, 3, 4)
```

**Types and Tests**

Given a vector, you can determine its class with `class`, or check if
it's a specific type with an "is" function: `is.character()`,
`is.numeric()`, `is.integer()`, `is.logical()`, or, more generally,
`is.atomic()`.

```{r }
char_var <- c("harry", "sally")

class(char_var)

is.character(char_var)

is.atomic(char_var)

num_var <- c(1, 2.5, 4.5)

class(num_var)

is.numeric(num_var)

is.atomic(num_var)
```

NB: `is.vector()` does not test if an object is a vector. Instead, it
returns `TRUE` only if the object is a vector with no attributes apart
from names. Use `is.atomic(x) || is.list(x)` to test if an object is
actually a vector.

**Coercion**

All atomic vector elements must be the same type, so when you combine
different types, they will be **coerced** to the **most flexible type.**
Types from least to most flexible are: logical \> integer \> double \>
character.

For example, combining a character and an integer yields a character:

```{r}
str(c("a", 1))
```

**Guess what the following do without running them first**

```{r, eval = FALSE}
c(1.7, "a")

c(TRUE, 2)

c("a", TRUE)
```

Notice that when a logical vector is coerced to an integer or double,
`TRUE` becomes 1, and `FALSE` becomes 0. This is very useful in
conjunction with `sum()` and `mean()`

```{r}
x <- c(FALSE, FALSE, TRUE)

as.numeric(x)

# Total number of TRUEs
sum(x)

# Proportion that is TRUE
mean(x)
```

Coercion often happens automatically. This is called implicit coercion.
Most mathematical functions (`+`, `log`, `abs`, etc.) will coerce to a
numeric or integer, and most logical operations (`&`, `|`, `any`, etc)
will coerce to a logical. You will usually get a warning message if the
coercion might lose information.

```{r}
1 < "2"

"1" > 2
```

You can also coerce vectors explicitly coerce with `as.character()`,
`as.numeric()`, `as.integer()`, or `as.logical()`. Example:

```{r}
x <- 0:6

as.numeric(x)

as.logical(x)

as.character(x)
```

Sometimes coercions, especially nonsensical ones, won't work.

```{r}
x <- c("a", "b", "c")

as.numeric(x)

as.logical(x)
```

**Short Exercise**

```{r, eval=FALSE}
# 1. Create a vector of a sequence of numbers between 1 to 10.
x <- 1:10 
# 2. Coerce that vector into a character vector

# 3. Add the element "11" to the end of the vector
x <- c(as.character(x), "11")
# 4. Coerce it back to a numeric vector.
as.numeric(x)
```

## Lists

Lists are also vectors, but different from atomic vectors because their
elements can be of any type. In short, they are generic vectors. For
example, you construct lists by using `list()` instead of `c()`:

Lists are sometimes called recursive vectors, because a list can contain
other lists. This makes them fundamentally different from atomic
vectors.

```{r}
x <- list(1, "a", TRUE, c(4, 5, 6))

x
```

You can coerce other objects using `as.list()`. You can test for a list
with `is.list()`

```{r}
x <- 1:10

x <- as.list(x)

is.list(x)

x <- list(x, c(2,3,4))
```

`c()` will combine several lists into one. If given a combination of
atomic vectors and lists, `c()` (con**c**atenate) will coerce the
vectors to lists before combining them. Compare the results of `list()`
and `c()`:

```{r}
x <- list(list(1, 2), c(3, 4))

y <- c(list(1, 2), c(3, 4))
```

You can turn a list into an atomic vector with `unlist()`. If the
elements of a list have different types, `unlist()` uses the same
coercion rules as `c()`.

```{r}
x <- list(list(1, 2), c(3, 4))
     
str(x)

unlist(x)
```

Lists are used to build up many of the more complicated data structures
in R. For example, both data frames and linear models objects (as
produced by `lm()`) are lists:

```{r}
is.list(mtcars)

mod <- lm(mpg ~ wt, data = mtcars)

is.list(mod)
```

For this reason, lists are handy inside functions. You can "staple"
together many different kinds of results into a single object that a
function can return.

A list does not print to the console like a vector. Instead, each
element of the list starts on a new line.

```{r}
x.vec <- c(1, 2, 3)
x.list <- list(1, 2, 3)
x.vec
x.list
```

For lists, elements are **indexed by double brackets**. Single brackets
will still return a(nother) list. (We'll talk more about subsetting and
indexing in the fourth lesson.)

```{r}
x <- list(c("a", "b"), c("c", "b"), "c")
```

**Exercises**

1.  What are the four basic types of atomic vectors? How does a list differ from an atomic vector?

2.  Why is `1 == "1"` true? Why is `-1 < FALSE` true? Why is `"one" < 2` false?

3.  Create three vectors and then combine them into a list.

4.  If `x` is a list, what is the class of `x[1]`? How about `x[[1]]`?

## Attributes {#attributes}

Attributes provide additional information about the data to you, the
user, and to R. We've already seen the following three attributes in
action:

-   Names (`names(x)`), a character vector giving each element a name.

-   Dimensions (`dim(x)`), used to turn vectors into matrices.

-   Class (`class(x)`), used to implement the S3 object system.

**Additional tips**

In an object-oriented system, a
[class](https://www.google.com/search?q=what+is+class+programming&oq=what+is+class+programming&aqs=chrome.0.0l6.3543j0j4&sourceid=chrome&ie=UTF-8)
(an extensible problem-code-template) defines a type of object like what
its properties are, how it behaves, and how it relates to other types of
objects. Therefore, technically, an object is an
[instance](https://en.wikipedia.org/wiki/Instance_(computer_science))
(or occurrence) of a class. A method is a function associated with a
particular type of object.

### Names

You can name a vector when you create it:

```{r}
x <- c(a = 1, b = 2, c = 3)
```

You can also modify an existing vector:

```{r}
x <- 1:3

names(x) <- c("e", "f", "g")
```

Names don't have to be unique. However, character subsetting, described in the next lesson, is the most important reason to use names, and it is
most useful when the names are unique. (For Python users, a vector behaves like a Python dictionary key when names are unique.)

Not all elements of a vector need to have a name. If some names are missing, `names()` will return an empty string for those elements. If all names are missing, `names()` will return `NULL`.

```{r}
y <- c(a = 1, 2, 3)

names(y)

z <- c(1, 2, 3)

names(z)
```

You can create a new vector without names using `unname(x)`, or remove
names in place with `names(x) <- NULL`.

## Factors

```{r}
col <- c("red", "blue", "yellow")

income <- c("low", "middle", "high")
```

Factors are special vectors that represent categorical data. Factors can
be ordered (ordinal variable) or unordered (nominal or categorical
variable) and are important for modeling functions such as `lm()` and
`glm()` and also in plot methods.

**Quiz**

1.  If you want to enter dummy variables (Democrats = 1, Non-democrats =
    0)  in your regression model, should you use a numeric or factor
        variable?

Factors can only contain pre-defined values. Set allowed values using
the `levels()` attribute. Note that a factor's levels will always be
character values.

```{r}
x <- c("a", "b", "b", "a")

x <- factor(c("a", "b", "b", "a"))

x

class(x)

levels(x)

# You can't use values that are not in the levels
x[2] <- "c"

# NB: you can't combine factors
c(factor("a"), factor("b"))

rep(1:5, rep(6, 5))
```

Factors are pretty much integers that have labels on them. Underneath,
it's really numbers (1, 2, 3...).

```{r}
x <- factor(c("a", "b", "b", "a"))

str(x)
```

They are better than using simple integer labels because factors are
what are called self-describing. For example, `democrat` and
`republican` is more descriptive than `1`s and `2`s.

Factors are useful when you know the possible values a variable may
take, even if you don't see all values in a given dataset. Using a
factor instead of a character vector makes it obvious when some groups
contain no observations:

```{r}
party_char <- c("democrat", "democrat", "democrat")

party_char

party_factor <- factor(party_char, levels = c("democrat", "republican"))

party_factor

table(party_char) # shows only democrats

table(party_factor) # shows republicans too
```

Sometimes factors can be left unordered. Example: `democrat`,
`republican.`

Other times you might want factors to be ordered (or ranked). Example:
`low`, `medium`, `high`.

```{r}
x <- factor(c("low", "medium", "high"))

x

is.ordered(x)

y <- ordered(c("low", "medium", "high"), levels = c("high", "medium", "low"))

is.ordered(y)

y
```

While factors look (and often behave) like character vectors, they are integers. So be careful when treating them like strings. Some string
methods (like `gsub()` and `grepl()`) will coerce factors to strings, while others (like `nchar()`) will throw an error, and still others
(like `c()`) will use the underlying integer values.

```{r}
x <- c("a", "b", "b", "a")

x

is.factor(x)

x <- as.factor(x)

x

c(x, "c")
```

For this reason, it's usually best to explicitly convert factors to
character vectors if you need string-like behavior. There was a memory
advantage to using factors instead of character vectors in early
versions of R, but this is no longer the case.

Unfortunately, most data loading functions in R automatically convert
character vectors to factors. This is suboptimal, because there's no way
for those functions to know the set of all possible levels or their
optimal order. If this becomes a problem, use the argument
`stringsAsFactors = FALSE` to suppress this behavior and manually
convert character vectors to factors using your knowledge of the data.

**More attributes**

All R objects can have arbitrary additional attributes used to store
metadata about the object. Attributes can be considered a named list
(with unique names). Attributes can be accessed individually with
`attr()` or all at once (as a list) with `attributes().`

```{r}
y <- 1:10

attr(y, "my_attribute") <- "This is a vector"

attr(y, "my_attribute")

# str returns a new object with modified information
str(attributes(y))
```

**Exercises**

1.  What happens to a factor when you modify its levels?

```{r, results = "none"}
f1 <- factor(letters)

levels(f1) <- rev(levels(f1))

f1
```

2.  What does this code do? How do `f2` and `f3` differ from `f1`?

```{r, results = "none"}
f2 <- rev(factor(letters))

f3 <- factor(letters, levels = rev(letters))
```
