---
title: 'Visualizing'
author: "Jae Yeon Kim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

# Setup 

```{r}
pacman::p_load(tidyverse, gapminder, nycflights13, ggthemes)
```

## Visualizing (ggplot2)

- The following material is adapted from Kieran Healy's excellent book (2019) on [data visualization](https://socviz.co/) and Hadley Wickham's equally excellent book on [ggplot2](https://ggplot2-book.org/). For more theoretical discussions, I recommend you to read [The Grammar of Graphics](https://link.springer.com/book/10.1007%2F0-387-28695-0) by Leland Wilkinson.

- Why should we care about data visualization? More precisely, why should we learn the grammar of statistical graphics?
- Sometimes, pictures are better tools than words in 1) exploring, 2) understanding, and 3) explaining data.

### Motivation 

[Anscombe](https://en.wikipedia.org/wiki/Frank_Anscombe)'s quarter comprises four datasets, which are so alike in terms of their descriptive statistics but quite different when presented graphically.

```{r}
# Set theme
theme_set(theme_minimal())
```

```{r}
# Data
anscombe
```

```{r}
# Correlation
cor(anscombe)[c(1:4), c(5:8)]
```

```{r}
# gather and select
anscombe_processed <- anscombe %>%
  gather(x_name, x_value, x1:x4) %>%
  gather(y_name, y_value, y1:y4)

# plot
anscombe_processed %>%
  ggplot(aes(x = x_value, y = y_value)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  facet_grid(x_name ~ y_name) +
  theme_bw() +
  labs(
    x = "X values",
    y = "Y values",
    title = "Anscombe's quartet"
  )
```

### The grammar of graphics 

- the grammar of graphics 

    - data
    - aesthetic attributes (color, shape, size)
    - geometric objects (points, lines, bars)
    - stats (summary stats)
    - scales (map values in the data space)
    - coord (data coordinates)
    - facet (facetting specifications)
    
No worries about new terms. We're going to learn them by actually plotting. 

- Workflow: 

    1. Tidy data 
    2. Mapping 
    3. Geom 
    4. Cor_ordinates and scales 
    5. Labels and guides
    6. Themes
    7. Save files 

### mapping and geom

- `aes` (aesthetic mappings or aesthetics) tells which variables (x, y) in your data should be represented by which visual elements (color, shape, size) in the plot.

- `geom_` tells the type of plot you are going to use 

### basic aes (x , y)

```{r}
p <- ggplot(
  data = gapminder,
  mapping = aes(x = gdpPercap, y = lifeExp)
) # ggplot or R in general takes positional arguments too. So, you don't need to name data, mapping each time you use ggplot2.

p

p + geom_point()

p + geom_point() + geom_smooth() # geom_smooth has calculated a smoothed line;
# the shaded area is the standard error for the line
```

### Univariate distribution

- `geom_histogram()`: For the probability distribution of a continuous variable. Bins divide the entire range of values into a series of intervals (see [the Wiki entry](https://en.wikipedia.org/wiki/Histogram)). 
- `geom_density()`: Also for the probability distribution of a continuous variable. It calculates a [kernel density estimate](https://en.wikipedia.org/wiki/Kernel_density_estimation) of the underlying distribution. 

#### Histogram 

```{r}
data(midwest) # load midwest dataset

midwest
```

```{r, eval = FALSE}
midwest %>%
  ggplot(aes(x = area)) +
  geom_point() # not working.
```

```{r}
midwest %>%
  ggplot(aes(x = area)) +
  geom_histogram() # stat_bin argument picks up 30 bins (or "bucket") by default.

midwest %>%
  ggplot(aes(x = area)) +
  geom_histogram(bins = 10) # only 10 bins.

ggplot(
  data = subset(midwest, state %in% c("OH", "IN")),
  mapping = aes(x = percollege, fill = state)
) +
  geom_histogram(alpha = 0.7, bins = 20) +
  scale_fill_viridis_d()
```

#### Density 

```{r}
midwest %>%
  ggplot(aes(x = area, fill = state, color = state)) +
  geom_density(alpha = 0.3) +
  scale_color_viridis_d() +
  scale_fill_viridis_d()
```

### Advanced aes (size, color)

- There's also `fill` argument (mostly used in `geom_bar()`). Color `aes` affects the appearance of lines and points, fill is for the filled areas of bars, polygons, and in some cases, the interior of a smoother's standard error ribbon.

- The property size/color/fill represents... 

```{r}
ggplot(
  data = gapminder,
  mapping = aes(
    x = gdpPercap, y = lifeExp,
    size = pop
  )
) +
  geom_point()
```

```{r}
ggplot(
  data = gapminder,
  mapping = aes(
    x = gdpPercap, y = lifeExp,
    size = pop,
    color = continent
  )
) +
  geom_point() +
  scale_color_viridis_d()
```

```{r}
# try red instead of "red"
ggplot(
  data = gapminder,
  mapping = aes(
    x = gdpPercap, y = lifeExp,
    size = pop,
    color = "red"
  )
) +
  geom_point()
```

Aesthetics also can be mapped per Geom. 

```{r}
p + geom_point() +
  geom_smooth()

p + geom_point(alpha = 0.3) + # alpha controls transparency
  geom_smooth(color = "red", se = FALSE, size = 2)

p + geom_point(alpha = 0.3) + # alpha controls transparency
  geom_smooth(color = "red", se = FALSE, size = 2, method = "lm")
```

```{r}
ggplot(
  data = gapminder,
  mapping = aes(
    x = gdpPercap, y = lifeExp,
    color = continent
  )
) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", color = "red") +
  labs(
    x = "log GDP",
    y = "Life Expectancy",
    title = "A Gapminder Plot",
    subtitle = "Data points are country-years",
    caption = "Source: Gapminder"
  )

ggplot(
  data = gapminder,
  mapping = aes(
    x = gdpPercap, y = lifeExp,
    color = continent,
    fill = continent
  )
) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", color = "red") +
  labs(
    x = "log GDP",
    y = "Life Expectancy",
    title = "A Gapminder Plot",
    subtitle = "Data points are country-years",
    caption = "Source: Gapminder"
  ) +
  scale_color_viridis_d() +
  scale_fill_viridis_d()
```

### Co-ordinates and scales 

```{r}
p + geom_point() +
  coord_flip() # coord_type
```

The data is heavily bunched up against the left side. 

```{r}
p + geom_point() # without scaling

p + geom_point() +
  scale_x_log10() # scales the axis of a plot to a log 10 basis

p + geom_point() +
  geom_smooth(method = "lm") +
  scale_x_log10()
```


### Labels and guides 

`scales` package has some useful premade formatting functions. You can either load scales or just grab the function you need from the library using `scales::` 

```{r}
p + geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", color = "red") +
  scale_x_log10(labels = scales::dollar) +
  labs(
    x = "log GDP",
    y = "Life Expectancy",
    title = "A Gapminder Plot",
    subtitle = "Data points are country-years",
    caption = "Source: Gapminder"
  )
```

### Themes

```{r}
p + geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", color = "red") +
  scale_x_log10(labels = scales::dollar) +
  labs(
    x = "log GDP",
    y = "Life Expectancy",
    title = "A Gapminder Plot",
    subtitle = "Data points are country-years",
    caption = "Source: Gapminder"
  ) +
  theme_economist()
```

### ggsave 

```{r eval = FALSE}
figure_example <- p + geom_point(alpha = 0.3) +
  geom_smooth(method = "gam", color = "red") +
  scale_x_log10(labels = scales::dollar) +
  labs(
    x = "log GDP",
    y = "Life Expectancy",
    title = "A Gapminder Plot",
    subtitle = "Data points are country-years",
    caption = "Source: Gapminder"
  ) +
  theme_economist()

ggsave(figure_example, here("outputs", "figure_example.png"))
```