---
title: "Assignment review"
subtitle: "Week 1"
author: "Jae Yeon Kim"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
---

# Overview

```{r}
# load pkg
pacman::p_load(here, dplyr)

# find the file path 
file_name <- list.files(here("reviews"))[grep("2022_01", list.files(here("reviews")))]

# assign the file to an object 
ass2 <- readxl::read_xlsx(here("reviews", file_name))

ass2 <- ass2 %>%
    filter(KorName != "김재연")
```

---

```{r}
score <- ass2 %>%
    rename(score = `Assignment: Reshaping data (Real)`) %>%
    mutate(score = as.numeric(score)) %>%
    pull(score)

h <- hist(score)

plot(h, freq = FALSE)
```

---

# Why do we have a bimodal distribution?

- The assignment was not difficult. 

- Three major problems explain the wide variation. Some of you did well, and others didn't. I'm sure most of you can do better in the next assignment.

- Note that the assignment counts 50% of the total evaluation (final exam counts 25%). 

- There will be three more assignments (in the same format, but they will be harder). 

- Your performance will be graded on an absolute basis.

---

# Pr 1: Read the instruction carefully. 

1. You need to submit the assignment using an RMD and an HTML rendered output (I addressed and emphasized this twice). Some of you only submitted the HTML file. The result is damaging. If you missed one of those files, you're likely to get the check minus, even if you did okay in the other areas.

2. You need to answer all questions (!). This seems obvious, but some of you did not answer the final question (what would be the ideal data type of the `df$valid_start` and `df$valid_end.`)

3. You need to provide your own reasoning for open-ended questions like the final question. Some of you provided the reasoning based on memory efficiency and others on workflow. If you didn't provide reasoning, you wouldn't get the check plus.

---

# Pr 2: Styles 

- Use markdown properly. 

`#, ##, ###` : these are section, subsection, and subsection headings. 

- Use functions properly. 

```{r eval = FALSE}
# good 
tidytuesdayR::tt_load()

# bad 
tidytuesdayR::tt_load # () missing 
library(tidytuesdayR, include.only = tt_load()) # too long 
```

---

Use spacing properly. 

```{r eval = FALSE}
# good 
target <- tt_load(2021, 30)

# bad
target<-tt_load(201, 30)
```

This one is really problematic.

```{r eval = FALSE}
# good 
class(target)

# bad 
class (target)
```

---

# Pr3: Substance 

```{r}
# Next time I will skip the tidytuesdayR pkg to avoid GitHub related issues 
df <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-07-20/drought.csv')
```

---

```{r}
nrow(df)

df$state_abb[1:5]

class(df$state_abb[1:5]) # don't print the whole thing!
```

---

```{r}
names(df)
# df[,2][1:5] (this doesn't work!)
class(df[1:5,2])
```

---

```{r}
class(df[,2]) # dataframe 

typeof(df[,2]) # heterogeneous and multiple dimensions 

str(df[,2]) # useful, but sometimes too much information
```

---

# Takeaways. 

1. Read instructions carefully. 

2. For open ended-questions, provide your reasoning. 

3. Style your code properly.